---
title: "Introduction to pama"
output: rmarkdown::html_vignette
description: >
  Start here if this is your first time using dplyr. You'll learn the basic   
  philosophy, the most important data manipulation verbs, and the pipe, `%>%`, 
  which allows you to combine multiple verbs together to solve real problems.
vignette: >
  %\VignetteIndexEntry{pama}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(pama)
```

When working on **long-life projects**, you must **ensure reproducibility** on new machines. For this, the following steps are fundamental:

-   Understand the basis for reproducibility (e.g. external connections, technologies such as R and its packages).
-   New R and/or package versions can change functions' inputs, outputs and functionality.
-   Therefore, be able to save and restore the versions of the package and of R with which it was built.
-   Allow easy access to different versions of the packages to facilitate updates

The "pama" package makes these requirements quick and easy to implement:

-   **Working on the project**
    -   Using `create_requirements()` a "requirements" file can be created that contains all necessary information (R version, package name & package version) to replicate the used packages in your R scripts.
    -   Loading package version lists from the "requirements" file is enabled using `load_requirements()`.
    -   Individual packages with specific versions can be loaded with `library_version()`.
-   **Replicating the project**
-   In the new project directory a library folder with the standard packages of the desired R version can be created with `setupLibrary()`.
-   After that, all packages with the used versions from the "requirements" files can be installed into the new library folder using `install_requirements()`.

This document introduces you to pama's basic set of tools, and shows you how to apply them to your project.

## Project: titanic

To explore the functionality of `pama`, we will use the project ["titanic-r"](https://github.com/ColinLeverger/titanic-r/) by Colin Leverger. In this project, a predictive model was built that answers the question: "What sorts of people were more likely to survive the Titanic disaster?" using passenger data (i.e. name, age, gender, socio-economic class, etc).

```{r}
# set working directory to where the .zip file will be located
setwd(dir = file.path(Sys.getenv("USERPROFILE"),"Desktop"))

# download a .zip file of the repository from the "Clone or download - Download ZIP" button on the GitHub repository of interest
download.file(url = "https://github.com/ColinLeverger/titanic-r/archive/refs/heads/master.zip", 
              destfile = "titanic-r.zip")

# unzip the .zip file
unzip(zipfile = "titanic-r.zip")

# set the working directory to be inside the newly unzipped GitHub repository of interest
setwd(dir = file.path(Sys.getenv("USERPROFILE"),"Desktop","titanic-r-master"))

# examine the contents
list.files()
```

## 1. Working on the project

```{r, eval=F}

# Install "pama" if not already installed and load it
if(!require("pama")) install.packages("pama")
library(pama)
```

`pama` aims to facilitate the way of working both (1.) in projects with packages, and (2.) reproducing projects. We start here with the work on an existing project.

### A. Creating a "requirements" file

In order to make your project reproducible a central location where all the information of the packages and their versions utilized in the project is very helpful. Using `create_requirements()` a "requirements"-file will be created that contains all necessary information (incl. R version, package name & version structured according to your scripts) to A.) replicate your library-folder with the exact same package versions (-> `install_requirements()`) and/or B.) switch the utilized versions of your packages (-> `apply_library_version()`).

```{r}
setwd(dir = file.path(Sys.getenv("USERPROFILE"),"Desktop","titanic-r-master"))

#Creating a req-file in "C:/Project" called "requirements.txt", without list structure
create_reqirements_file(draft = FALSE, script.paths = "./", req.file.name = "requirements.txt")
#User inputs (randomly chosen):
  # dataQualityR: 1.0
  # dplyr: 1.1.2
  # ggplot2: 3.4.2
  # reshape2: 1.4.4
  # vcd: 1.4-11
  # randomForest: 4.7-1.1
```

...

## 2. Replicating the project

In the new project directory a library folder with the standard packages of the desired R version can be created with `setupLibrary()`.

```{r}
setwd(dir = file.path(Sys.getenv("USERPROFILE"),"Desktop","titanic-r-master"))

#Creating a "lib"-folder with all standard packages of the specified R Version (-> specified by R.home())
setupLibrary(new.lib = "lib", main.lib.path = paste0(R.home(), "/library"))
```

After that, all packages with the used versions from the "requirements" files can be installed into the new library folder using `install_requirements()`.


```{r, results='hold'}
setwd(dir = file.path(Sys.getenv("USERPROFILE"),"Desktop","titanic-r-master"))

#Install all package versions specified in "requirements.txt"-file in the WD
install_requirements()
```


If your R-scripts still contain the "old" `library()`- / `requires()`-syntax you can automatically replace it using `apply_library_version()`.
```{r}
setwd(dir = file.path(Sys.getenv("USERPROFILE"),"Desktop","titanic-r-master"))

#Adjust package-loading in R scripts to pama-style
  #library(<package>,) & require(<package>,) -> library_version(<package>, <version>)
apply_library_version(script.paths = "./Scripts")
```



...


When publishing your project, you can cite all utilized packages using `citations2()`:

```{r}
setwd(dir = file.path(Sys.getenv("USERPROFILE"),"Desktop","titanic-r-master"))

#Creating a lib_citations_<date>.csv from the "requirements.txt" within the working directory
citations2()
```
Introduction is still being expanded...
