---
title: "Introduction to pama"
output: rmarkdown::html_vignette
description: >
  Start here if this is your first time using dplyr. You'll learn the basic   
  philosophy, the most important data manipulation verbs, and the pipe, `%>%`, 
  which allows you to combine multiple verbs together to solve real problems.
vignette: >
  %\VignetteIndexEntry{pama}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

When working on **long-life projects**, you must **ensure reproducibility** on new machines. For this, the following steps are fundamental:

-   Understand the basis for reproducibility (e.g. external connections, technologies such as R and its packages).
-   New R and/or package versions can change functions' inputs, outputs and functionality.
-   Therefore, be able to save and restore the versions of the package and of R with which it was built.
-   Allow easy access to different versions of the packages to facilitate updates

The "pama" package makes these requirements quick and easy to implement:

-   **Working on the project**
    -   Using `create_requirements()` a "requirements" file can be created that contains all necessary information (R version, package name & package version) to replicate the used packages in your R scripts.
    -   Loading package version lists from the "requirements" file is enabled using `load_requirements()`.
    -   Individual packages with specific versions can be loaded with `library_version()`.
-   **Replicating the project**
-   In the new project directory a library folder with the standard packages of the desired R version can be created with `setupLibrary()`.
-   After that, all packages with the used versions from the "requirements" files can be installed into the new library folder using `install_requirements()`.

This document introduces you to pama's basic set of tools, and shows you how to apply them to your project.


## Install and load pama

```{r, eval = FALSE}
# Install from GitHub (I)
  if(!require("remotes")) install.packages("remotes")
  library("remotes")

  install_github("https://github.com/SimonRess/PaMaTo/", build_manual = TRUE, build_vignettes = TRUE) # also get the vignette
  
  library("pama")
  

  
  # OR

  
# Install from GitHub (II)
  if(!require("devtools")) install.packages("devtools")
  library(devtools)
  
  install_github("https://github.com/SimonRess/PaMaTo/", build_manual = TRUE, build_vignettes = TRUE) # also get the vignette
  
  library("pama")
```


Take a loot at this Vignette locally
```{r, eval = FALSE}
help(package = "pama")

#then choose: 
```



---

## Project: titanic

To explore the functionality of `pama`, we will use the project ["titanic-r"](https://github.com/ColinLeverger/titanic-r/) by Colin Leverger. In this project, a predictive model was built that answers the question: "What sorts of people were more likely to survive the Titanic disaster?" using passenger data (i.e. name, age, gender, socio-economic class, etc).

```{r, eval=FALSE}
# set working directory to where the .zip file will be located
setwd(dir = file.path(Sys.getenv("USERPROFILE"),"Desktop"))

# download a .zip file of the repository from the "Clone or download - Download ZIP" button on the GitHub repository of interest
download.file(url = "https://github.com/ColinLeverger/titanic-r/archive/refs/heads/master.zip", 
              destfile = "titanic-r.zip")

# unzip the .zip file
unzip(zipfile = "titanic-r.zip")

# set the working directory to be inside the newly unzipped GitHub repository of interest
setwd(dir = file.path(Sys.getenv("USERPROFILE"),"Desktop","titanic-r-master"))

# examine the contents
list.files()
```


---


## 1. Make your project reproducible

**Content**
- 1.1 `create_reqirements_file()` function
- 1.2 `apply_library_version()` function


```{r, eval=F}

# Install "pama" if not already installed and load it
if(!require("pama")) install.packages("pama")
library(pama)
```

`pama` aims to facilitate the way of working both (1.) in projects with packages, and (2.) reproducing projects. We start here with the work on an existing project.


### 1.1 Creating a "requirements" file: `create_requirements()`

In order to make your project reproducible a central location where all the information of the packages and their versions utilized in the project is very helpful. Using `create_requirements()` a "requirements"-file will be created that contains all necessary information (incl. R version, package name & version structured according to your scripts) to A.) replicate your library-folder with the exact same package versions (-> `install_requirements()`) and/or B.) switch the utilized versions of your packages (-> `apply_library_version()`).

```{r, eval=FALSE}
setwd(dir = file.path(Sys.getenv("USERPROFILE"),"Desktop","titanic-r-master"))

#Creating a req-file in "C:/Project" called "requirements_<date>.txt", with list structure, but no package versions
create_reqirements_file(draft = FALSE, script.paths = "./")
#User inputs (randomly chosen):
  # dataQualityR: 1.0
  # dplyr: 1.1.2
  # ggplot2: 3.4.2
  # reshape2: 1.4.4
  # vcd: 1.4-11
  # randomForest: 4.7-1.1


#If there's a lib folder: Creating a req-file in "C:/Project" called "requirements.txt", with list structure, including package versions
#create_reqirements_file(draft = FALSE, script.paths = "./", req.file.name = "requirements.txt", lib.path = <....>)
```


### 1.2 Adjust package-loading in R scripts to pama-style: `apply_library_version()`
In order to ensure that always the correct versions of the packages used are loaded in the future, they must be specified in the code. `apply_library_version()` replaces all functions defined in `<patterns>` (default: `library()` and `require()`) by the pama-function
`library_version(<package_name>, <package_version>)`. The `<package_name>` attribute is extracted from the scripts. The `<package_version>` is read from the specified requirements-file.


```{r, eval=FALSE}
setwd(dir = file.path(Sys.getenv("USERPROFILE"),"Desktop","titanic-r-master"))

#Adjust package-loading within R scripts in .\Scripts to pama-style, using versions from the "requirements.txt"-file
apply_library_version(req.file.path = r"(C:\Users\sress\Desktop\titanic-r-master)",
                      req.file.name = paste0("requirements_",Sys.Date(), ".txt"),
                      script.paths = r"(C:\Users\sress\Desktop\titanic-r-master\Scripts)")
```

---


## 2. Reproducing your project

**Content**
-   2.1 `setupLibrary()` function
-   2.2 `install_requirements()` function


### 2.1 Setup a project library: `setupLibrary()`

In the new project directory a library folder with the standard packages of the desired R version can be created with `setupLibrary()`.

```{r, eval=FALSE}
setwd(dir = file.path(Sys.getenv("USERPROFILE"),"Desktop","titanic-r-master"))

#Creating a "lib"-folder with all standard packages of the specified R Version (-> here specified by R.home())
setupLibrary(new.lib = "lib", main.lib.path = paste0(R.home(), "/library"))
```


### 2.2 Installing requires packages: `install_requirements()`

After that, all packages with the used versions from the "requirements" files can be installed into the new library folder using `install_requirements()`.


```{r, eval=FALSE}
setwd(dir = file.path(Sys.getenv("USERPROFILE"),"Desktop","titanic-r-master"))

#Install all package versions specified in "requirements_<date>.txt"-file in the WD
install_requirements(req.file.name = paste0("requirements_",Sys.Date(), ".txt"))
```


---


## 3. Work on your project

-   3.1 `modify_libPaths()` function
-   3.2 `install_package_version()` function
-   3.3 `library_version()` function
-   3.4 `load_requirements()` function
-   3.5 `citations2()` function


### 3.1 `modify_libPaths()` function
...

```{r, eval=FALSE}
setwd(dir = file.path(Sys.getenv("USERPROFILE"),"Desktop","titanic-r-master"))

modify_libPaths(reset = file.path(getwd(),"lib"))
```


### 3.2 `install_package_version()` function

While working on your project there's sometimes the need to install a new package or another version. In order to keep your project reproducible you should already in your code specify which version of a package you are installing. This can be done by using `install_package_version()` which should replace the R Base function `install.packages()`:

```{r, eval=FALSE}
setwd(dir = file.path(Sys.getenv("USERPROFILE"),"Desktop","titanic-r-master"))

#Within your script simply use install_package_version() instead of install.packages()

  #Install ggplot2 version 3.0.0 into your projects lib-folder
  install_package_version("ggplot2", "3.0.0", lib.install.path = file.path(getwd(),"lib"))
  
  #if your packages library path is already specified as .libPaths[1] you don't need to specify "lib.install.path"
  #-> this is achieved resetting the .libPaths(): modify_libPaths(reset = file.path(getwd(),"lib"))
  install_package_version("ggplot2", "3.0.0")
```


### 3.2 `library_version()` function
...
```{r, eval=FALSE}
setwd(dir = file.path(Sys.getenv("USERPROFILE"),"Desktop","titanic-r-master"))

# Load v3.4.0 of ggplot2 from folder "ggplot2_3.4.0" in search paths
library_version("ggplot2", "3.4.2")
```

### 3.3 `load_requirements()` function
...

```{r, eval=FALSE}
setwd(dir = file.path(Sys.getenv("USERPROFILE"),"Desktop","titanic-r-master"))

detach_none_base() # detach all none base packages
load_requirements(req.file.name = paste0("requirements_",Sys.Date(), ".txt"))
as.vector(apply(sapply(sessionInfo()$otherPkgs, \(x) x[c("Package", "Version")]), 2, \(x) paste(x, collapse = "_"))) # currently loaded packages incl. version

```


### 3.4 `citations2()` function

When publishing your project, you can cite all utilized packages using `citations2()`:

```{r, eval=FALSE}
setwd(dir = file.path(Sys.getenv("USERPROFILE"),"Desktop","titanic-r-master"))

#Creating a lib_citations_<date>.csv from the "requirements.txt" within the working directory
citations2(req.file.name = paste0("requirements_",Sys.Date(), ".txt"))
```


## 4. Useful helper functions

-   4.1 `detach_none_base()` function
-   4.2 `find_package_version_on_cran()` function
-   4.3 `get_dependencies()` function
-   4.4 `get_installed_packages()` function
-   4.5 `get_requirements()` function


Introduction is still being expanded...
